### 2014-01-24 brains-introduction.tex
Rework "Related work" a little

The relevant literature on \emph{public representation dynamics} features two main streams. On one hand, we find studies aiming at understanding the macroscopic phenomena of \emph{social diffusion} of public representations, describing for instance the propagation of cultural items across social networks such as blogspace \citep{Gruhl04}, the characteristic times and diffusion cycles both within these social networks and with respect to the topical dynamics of news media \citep{Leskovec09}, or the reciprocal influence between the social network topology and the distribution of issues \citep{Cointet09}.     
%L'état de l'art en ce qui concerne l'épidémiologie culturelle peut être résumé ainsi : d'un côté, des études %s'intégrant au deuxième groupe de disciplines
% permettent une bonne compréhension des phénomènes macroscopiques de \emph{diffusion sociale} des représentations publiques, tels que le chemin emprunté par une information atomique donnée, les temps caractéristiques et les cycles de diffusion dans ces réseaux, la relation entre blogs et sites de médias, ou encore l'autorité et l'influence exercée par certains n\oe{}uds du réseau. 
These studies are relatively independent from anthropology and cognition and are at the interface between data mining, complex systems and quantitative sociology (first and foremost social network analysis). Without necessarily relying on specific social science theories, %these studies are interesting in that they aim at describing cultural dynamics by using large social media corpuses. 
this research stream is of interest for its aim of cultural dynamics description by using large social media corpora.
% ,    Ces études sont relativement indépendantes de l'anthropologie et de la cognition et  se situent à l'interface entre le \emph{data mining}, l'étude des systèmes complexes, et la sociologie quantitative (e.g. sociologie des réseaux), en exploitant de grands corpus de données produits par des collectes sur Internet. Sans nécessairement s'appuyer sur des cadres théoriques précis ou intégrés, ces études présentent néanmoins la particularité de vouloir décrire les dynamiques culturelles en exploitant les possibilités ouvertes par les larges corpus de données issus d'Internet. %, allant de l'étude de l'impact des réseaux sociaux sur la diffusion des informations sur ces mêmes réseaux \citep{Bakshy12} à l'analyse du rôle de la presse en ligne dans la vie politique locale \citep{Parasie12}, en passant par le détail des interactions entre éditeurs sur Wikipedia \citep{Jurgens12} ou l'évolution du vocabulaire des langues au cours du temps \citep{Lieberman07}. 
%On peut citer ici l'étude de la propagation et de la diffusion de l'information dans la blogosphère faite par \citet{Gruhl04}, l'étude du \emph{news cycle} faite par \citet{Leskovec09} qui fait intervenir la plupart de ces aspects, ou encore l'analyse des influences réciproques entre un réseau social et le réseau sémantique formé par les contenus diffusés sur ce premier réseau, faite par \citet{Cointet09} ; %on l'a déjà mentionné, 
 %, ces thèmes étant explorés principalement par le premier groupe de disciplines.

On the other hand, the study of the \emph{transformation} of public representations has emerged only recently. For one, the notion of ``cultural attractor'' has been the focus of models rendering the evolution of some representations and behaviors not until a few years ago \citep{Claidiere07}. Among the empirical approaches on the mutation of representations, some of the most relevant studies to date consist of a series of papers investigating \emph{quotation} transformations from a large corpus of US blog posts, initially collected and studied by \citet{Leskovec09} and further analyzed by \citet{Simmons11} and \citet{omod-mult}. They generally show several types of regularities and propose models of diffusion-transformation of quotations, which may nonetheless appear to be relatively simplistic from a cognitive viewpoint.  One of the main conclusions of these works lies in the fact that even for the supposedly most stable type of public representation, namely quotations, it is indeed still possible to observe and measure significant transformations.  However, these studies address transformations by focusing on the properties of the author (e.g. news vs. blogs) of the quotation, or the surrounding public space (e.g. quote frequency in the corpus), rather than the very cognitive features which may trigger or, at least, bias these transformations. % Cependant ces auteurs n'étudient les transformations observées qu'en fonction du type d'auteur qui reproduit les citations, et pas en fonction des citations elles-mêmes (en particulier les questions d'ordre cognitif concernant la raison pour laquelle ces transformations ont lieu, ou la raison pour laquelle elles ont lieu dans tel sens et pas dans tel autre, ne sont pas abordées). %Les possibilités offertes par cette étude étant vastes, on a décidé de travailler sur les mêmes données que \citeauthor{Simmons11} pour traiter notre question.

At this level, we have to turn to the broader psycholinguistic literature which provides one of the main cognitive foundations of public representation evolution by studying the influence of word features on the ease of recall. This state-of-the-art is relatively developed and shows the positive impact of classical psycholinguistic variables, such as word frequency (see \citet{Yonelinas02} for a review), age-of-acquisition \citep{Zevin02}, number of phonemes and number of syllables (see for instance \citet{Rey98} and \citet{nick-diss}).

The study of network-based properties, while heavily discussed around the notion of connectionism and its normative processual models, appeared much more recently as an empirical investigation field \CN. Let us mention two interesting and recent studies on that matter, which demonstrate in a strictly \emph{in vitro} framework and at the vocabulary level that word properties computed on a word network are important factors in the cognitive processes and reproduction of those words. %Les deux autres travaux viennent de la psycholinguistique, et posent les fondations pour l'étude des propriétés des représentations (qui seront ici des citations). 
First, \citet{Griffiths07} analyze a task where patients are asked to speak out the first word which comes to their mind when they are presented with a random alphabet letter. The authors show that there exists a link between the ease of recall of words and one of their semantic features, namely their authority position (PageRank) in a language-wide semantic network built from external word association data. % property concernant une tâche où l'on présente aux sujets une lettre de l'alphabet en leur demandant de dire le premier mot commençant par cette lettre qui leur vient à l'esprit. Les auteurs montrent qu'une certaine propriété des mots (le PageRank calculé sur un réseau sémantique construit à partir de données d'associations de mots) est un bon facteur prédictif de quel mot sera rappelé parmi les mots commençant par la lettre demandée ; ils montrent ainsi un lien entre une caractéristique sémantique quantitative des mots et la facilité de rappel de ces mots dans cette tâche. 
A second psycholinguistic study by \citet{Chan10} shows, in a picture-naming task, that words come quicker when they have a higher clustering coefficient in an underlying phonological network (which, again, is  defined from external phonological data). %, étudie une tâche où l'on demande aux sujets de nommer oralement une image. Les auteurs montrent qu'une autre propriété des mots, phonologique cette fois (le coefficient de regroupement des mots calculé sur le réseau phonologique formé par ces mots), a un effet sur la rapidité à nommer l'image présentée : c'est un autre cas de lien entre une caractéristique quantitative (phonologique) d'un mot et la facilité à produire ce mot.

On the whole, the current psycholinguistic state-of-the-art seems to hint at two antagonistic types of results.  On one hand, part of the literature tends to show that recall is easier for the least ``awkward'' words; those whose age of acquisition is earlier, length is smaller, semantic network position is more central -- this is particularly true in tasks where participants are asked to form spontaneous associations or utter a word in response to a given signal \CN. On the other hand, when the task consists in remembering a specific list of items, ``awkward'' words are actually more easily remembered, possibly as they are more informative and plausibly more discernable \CN. The jury is still out as to whether reformulation alteration, i.e. spontaneous replacement of words when asked to repeat a given utterance, is rather of the former or latter sort.  Our paper additionally sheds light on this debate.

\smallskip
More broadly, we thus observe a gap between, on one side, macro-level empirical studies on diffusion dynamics in a social system and, on the other side, studies essentially focused on micro-level transformations of representations --- these latter studies being either strongly normative, or yielding information whose integration into realistic cultural epidemiology models remains difficult. %On remarque de plus que le volet cognitif, central pour les phénomènes étudiés, est relativement peu développé dans les modèles et peu exploré de façon empirique.
%Les deux dernières études, issues de la psycholinguistique et dans un cadre strictement \emph{in vitro} et strictly at the vocabulary level, montrent donc que des propriétés des mots du vocabulaire, calculées sur un réseau formé par ces mêmes mots, sont des facteurs importants dans la façon dont on les traite cognitivement et dans la façon dont on les produit. C'est justement l'élément qui nous manquait : en se basant sur ces trois travaux, on va pouvoir utiliser des propriétés ainsi calculées sur les mots des citations pour mesurer quantitativement les transformations subies par ces citations lorsqu'elles sont reproduites. Pour rendre l'analyse possible, on se concentrera sur des cas de \emph{substitution} d'un mot par un autre lorsqu'un auteur reproduit une citation : à chaque substitution observée, on assiste à un échange entre deux mots pour lesquels on va pouvoir comparer les propriétés sémantiques.

%%% GARBAGE

%Cependant, depuis plus d'une décennie le paysage des choses observables dans le domaine des représentations publiques est en train de changer radicalement, du fait du déluge de données provenant des interactions en lignes, permettant l'étude empirique des représentations dans de cadre de l'épidémiologie culturelle.

%Nous vivons aujourd'hui non pas avec un manque de données concernant les représentations publiques, mais au contraire sous un véritable déluge de telles données. Quel est ce déluge ? Il est le produit d'une collecte qui est née avec l'utilisation commerciale et massive d'Internet. Cette collecte n'est pas un phénomène mineur, ni même un phénomène d'ampleur moyenne : les services sur Internet enregistrent aujourd'hui la quasi-totalité de ce qui est mesurable des interactions entre les personnes connectées ainsi que des interactions des personnes avec les machines. Vu l'importance que ce réseau a pris dans nos vies de tous les jours, on assiste bien à une explosion de ces données ; à tel point que certains observateurs d'Internet estiment que 90\% des données stockées aujourd'hui par les services sur Internet ont étés générées il y a moins de deux ans.
%Sans être des enregistrements des interactions \emph{physiques} entre les personnes (``real life''), il s'agit bien d'observations sur le terrain nouveau que constituent l'action, la perception, et l'interaction connectées, au travers des machines et d'Internet. En un mot, elles sont en fait des observations en grandes quantités de \emph{représentations publiques}. %\footnote{Bien sûr, une bonne partie des données collectées sur Internet appartient à des entreprises qui n'ont aucun intérêt à les partager ni à les exploiter autrement que financièrement. Mais les données publiquement accessibles forment tout de même une base considérable, qu'il est tout à fait possible d'exploiter comme telle.}.
%L'approche qu'on a esquissée ici n'est qu'une des façons d'aborder l'étude de la culture et de la cognition de façon empirique. En effet, les courants de recherche qui gravitent autour de ce thème sont nombreux et variés ; voici une rapide présentation des champs concernés et de la façon dont ils interagissent.

%Un groupe de disciplines est particulièrement pertinent ici, à l'interface entre le \emph{data mining}, l'étude des systèmes complexes, et la sociologie quantitative (e.g. sociologie des réseaux). Un grand nombre de travaux apparaissent exploitant de grands corpus de données produits par des collectes sur Internet, allant de l'étude de l'impact des réseaux sociaux sur la diffusion des informations sur ces mêmes réseaux \citep{Bakshy12} à l'analyse du rôle de la presse en ligne dans la vie politique locale \citep{Parasie12}, en passant par le détail des interactions entre éditeurs sur Wikipedia \citep{Jurgens12} ou l'évolution du vocabulaire des langues au cours du temps \citep{Lieberman07}. Une partie de ces travaux n'inscrivent leurs résultats que dans des cadres théoriques naïfs (autour de thèmes pour lesquels il existe pourtant des théories réfléchies développées par des champs des sciences sociales), et relèvent parfois plus des sciences de l'informatique que de l'étude de la culture (d'aucuns parlent d'ailleurs de \emph{social computing}). Ce groupe de disciplines présente néanmoins la particularité de vouloir étudier la culture en exploitant les possibilités ouvertes par les larges corpus de données issus d'Internet.

%% ----------------------------------------
%\subsection{Relevant fields} % -----
%% ----------------------------------------
%
%L'idée d'étudier la culture en lien avec la cognition apparaît comme une extension naturelle pour plusieurs disciplines historiquement séparées qui, chacune de leur côté, s'en sont approchées en partant de leur propre espace de concepts et en étendant leurs outils. Les dialogues entre ces disciplines ne sont pas rares, mais les traditions et les épistémologies différentes ont rendu les unifications plus ardues. On peut distinguer au moins deux grands groupes de disciplines débattant autour de thématiques reliées à ce sujet.%\footnote{On regroupe ici les disciplines par les thèmes qu'elles débattent, plutôt que par les similarités entre leurs façons de traiter ces thèmes.}.
%
%Le premier groupe est issu des débats entre la biologie évolutionniste, la psychologie cognitive et les sciences sociales (et plus particulièrement l'anthropologie), desquels ont émergé les domaines de la psychologie évolutionniste et de l'anthropologie cognitive. %Plusieurs théories ont été avancées et sont débattues en parallèle par ces disciplines. On se doit de mentionner ici en premier lieu le débat autour du programme mémétique initié par \citet{Dawkins76} et pour lequel l'ouvrage collectif de \citet{Aunger00} est un bon exemple. Un deuxième terrain exploré par cet ensemble de disciplines est le développement de modèles évolutionnistes basés sur les normes (voir par exemple \citealp{Ehrlich05}), avec comme travail fondamental celui de \citet{Boyd85}. Le troisième champ, particulièrement important aujourd'hui, est précisément celui de l'épidémiologie culturelle, initié par \citet{Sperber96} et défendu face aux autres courants dans des travaux comme celui de \citet{Atran03}.
%
%Le deuxième groupe, plus diffus, provient des échanges entre le \emph{data mining}, l'étude des systèmes complexes, et la sociologie quantitative (e.g. sociologie des réseaux). Un grand nombre de travaux apparaissent exploitant de grands corpus de données produits par des collectes sur Internet, allant de l'étude de l'impact des réseaux sociaux sur la diffusion des informations sur ces mêmes réseaux \citep{Bakshy12} à l'analyse du rôle de la presse en ligne dans la vie politique locale \citep{Parasie12}, en passant par le détail des interactions entre éditeurs sur Wikipedia \citep{Jurgens12} ou l'évolution du vocabulaire des langues au cours du temps \citep{Lieberman07}. Une partie de ces travaux n'inscrivent leurs résultats que dans des cadres théoriques naïfs (autour de thèmes pour lesquels il existe pourtant des théories réfléchies développées par des champs des sciences sociales), et relèvent parfois plus des sciences de l'informatique que de l'étude de la culture (d'aucuns parlent d'ailleurs de \emph{social computing}). Ce groupe de disciplines présente néanmoins la particularité de vouloir étudier la culture en exploitant les possibilités ouvertes par les larges corpus de données issus d'Internet.

>>>>

The relevant literature on \emph{public representation dynamics} features two main streams.
On one hand, we find studies of the macroscopic \emph{social diffusion} of public representations, describing for instance the propagation of cultural artefacts across social networks such as blogspace \citep{Gruhl04}, the characteristic times and diffusion cycles both within these social networks and with respect to the topical dynamics of news media \citep{Leskovec09}, or the reciprocal influence between the social network topology and the distribution of issues \citep{Cointet09}.
These studies are relatively independent from anthropology and cognition and are at the interface between data mining, complex systems and quantitative sociology (first and foremost social network analysis).
Without necessarily relying on specific social science theories, this research stream is of interest for its use of large social media corpora in studying cultural dynamics.

On the other hand, the study of the \emph{transformation} of public representations has emerged only recently.
For one, models involving evolution and representations to study the notion of ``cultural attractor'' have appeared only a few years ago \citep{Claidiere07}.
Among the empirical approaches on the mutation of representations, some of the most relevant studies to date consist in a series of papers investigating \emph{quotation} transformations in a large corpus of US blog posts, initially collected and studied by \citet{Leskovec09} and further analyzed by \citet{Simmons11} and \citet{omod-mult}.
They show several types of regularities and propose diffusion-transformation models of the evolution of quotations, which may nonetheless appear to be relatively simplistic from a cognitive viewpoint.
One of the main conclusions of these works is that even for quotations, a type of public representation that should be among the most stable, it is still possible to observe and measure significant transformations.
However, these studies address transformations by focusing on the properties of the source of the quotation (e.g. news outlet v. blog), or the surrounding public space (e.g. quotation frequency in the corpus), rather than the very cognitive-level features which may determine or, at least, influence these transformations.

At this level, we have to turn to the broader psycholinguistic literature which provides one of the main cognitive foundations for public representation evolution by studying the influence of word features on the ease of recall.
This field is well developed and details the impact that classical psycholinguistic variables such as word frequency (see \citet{Yonelinas02} for a review), age-of-acquisition \citep{Zevin02}, number of phonemes or number of syllables (see for instance \citet{Rey98} and \citet{nick-diss}), have in this type of task.

Less classical linguistic variables, based on the study of semantic network properties, have recently appeared as an empirical investigation field, after having been heavily discussed around the notion of connectionism and its normative processual models (see for instance \citet{collins1975spreading}).
Let us mention two interesting and recent studies on that matter, which demonstrate in a strictly \emph{in vitro} framework and at the vocabulary level that word properties computed on a word network are important factors for the cognitive processes and reproduction of those words.
First, \citet{Griffiths07} analyse a task where patients are asked to name the first word which comes to their mind when they are presented with a random letter from the Latin alphabet.
The authors show that there exists a link between the ease of recall of words and one of their semantic features, namely their authority position (PageRank) in a language-wide semantic network built from external word association data.
A second psycholinguistic study by \citet{Chan10} shows, in a picture-naming task, that words come quicker when they have a higher clustering coefficient in an underlying phonological network (which, again, is  defined from external phonological data).

\rk{add \url{https://www.sciencedirect.com/science/article/pii/S0893608012000330}, \url{http://research.clps.brown.edu/austerweil/pdfs/papers/randomWalkNips2012.pdf}, \url{http://link.springer.com/article/10.3758/s13421-013-0312-y} to that picture}

\TB{On the whole, the current psycholinguistic state-of-the-art seems to hint towards two antagonistic types of results.
On one hand, part of the literature tends to show that recall is easier for the least ``awkward'' words; those whose age of acquisition is earlier, length is smaller, semantic network position is more central -- this is particularly true in tasks where participants are asked to form spontaneous associations or utter a word in response to a given signal \CN.
On the other hand, when the task consists in remembering a specific list of items, ``awkward'' words are actually more easily remembered, possibly as they are more informative and plausibly more discernible \CN.
The jury is still out as to whether reformulation alteration, i.e. spontaneous replacement of words when asked to repeat a given utterance, is rather of the former or latter sort.
Our paper additionally sheds light on this debate.}
\rk{But do we really care about this?}

\smallskip
Stepping back, we observe a gap between, on one side, macro-level empirical studies of the diffusion dynamics in a social system and, on the other side, studies focused on micro-level transformations of representations --- these latter studies being either strongly normative, or whose results are difficult to integrate with realistic cultural epidemiology models.
